# **RStudio Exercise 6: Analysis of Longitudinal Data**

\  
```{r include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(psych)
```
First we will open all the required data files. (For detailed documentation how the data was created see: [https://github.com/ohwhalme/IODS-project/blob/master/data/meet_and_repeat.R](https://github.com/ohwhalme/IODS-project/blob/master/data/meet_and_repeat.R))
```{r}
BPRS <- read.csv("C:/Users/OHW/Desktop/R/IODS-project/data/BPRS.csv")
BPRSL <- read.csv("C:/Users/OHW/Desktop/R/IODS-project/data/BPRSL.csv")
RATS <- read.csv("C:/Users/OHW/Desktop/R/IODS-project/data/RATS.csv")
RATSL <- read.csv("C:/Users/OHW/Desktop/R/IODS-project/data/RATSL.csv")
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
```
Then we will proceed to "Implement the analyses of Chapter 8 of MABS using the RATS data".
\  

\  

#### **Comparing long and wide form of RATS**:
```{r}
glimpse(RATS)
head(RATS)
tail(RATS)
glimpse(RATSL)
head(RATSL)
tail(RATSL)
```
The wide form consist of 16 observations and 13 variables and the long form of 176 observations and 5 variables. As explained earlier (see above the link for how the data was created), in the data set we have 16 rats who have been assigned to 3 different experimental groups (diet) and their weight have been followed for 9 weeks. Wide form represents each individual on single row and the measurement points (WD1 to WD64) as separate variables. In the long form each participant is stacked on top of each other 11 times (reflecting the number of measurement points) and each days outcome (grams) are recoded into single variable.
\  

\  

#### **Presenting RATS visually and standardizing the data**:
```{r}
p1 <- ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p6
```
\  

Above we have plotted the 8 rats from group 1, 4 rats from group 2, and 4 rats from group 3 across the 9 week time period. On y-axis we have assigned rats weight on grams. We can infer that the group 1 is considerably lighter from base line to end of the trial than the groups 2 and 3 (i.e. rats were smaller to begin with in group 1). There seems to exist slight upwards trend in all of the groups indicating that the rats are gaining weight as the trial goes on. There does not seem to be great differences in the variability of individual rats weight across time, although the intercepts do differ quite much (across groups, but also within groups we can see at least one individual whos starting weight is considerably higher/lower that others in the group).
\  

Next, we will standardize the outcome values (grouped by time: Weight - mean(Weight))/sd(Weight)) and create the same plot.
```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)

p1 <- ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(name = "standardized Weight")
p6
```

Now "Weight" has mean of 0 and standard deviation of 1 on each measurement point (time). This highlights the fact that different individuals have different intercepts on "Weight". Because the standardation was grouped by time, the trends look somewhat different when compared to the first plot (standardization was mainly done to better show the "tracking" phenomenon, not so much to interpret the results).
\  

Next we will plot mean response profiles for the 3 dietary groups.
```{r}
n <- RATSL$Time %>% unique() %>% length()

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)

p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = c(1:3))
p3 <- p2 + geom_point(size=3) + scale_shape_manual(values = c(1:3))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + theme(legend.position = c(0.95,0.5))
p7 <- p6 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p7
```
\  

On x-axis we have plotted "Time" (i.e. measurement points) and on y-axis we have weight in grams. Different lines represent different experimental groups. Each score indicate particular groups mean weight in particular measurement point (standard errors are included as error bars). We can infer from the plot that, when mean response scores are scrutinized, different groups have different intercepts. Baseline for group 1 is considerably lower than baselines for groups 2 or 3. Similarly, baseline for group 2 is slightly lower when compared to group 3. Error bars indicate that variation is greatest in group 2 and lowest in group 1 (which most likely reflects the difference in sample size between the group 1 and the other groups). The slopes indicate roughly similar trend in all of the groups, so it's difficult to infer whether the experimental manipulation is having any effect.
\  

Next, to investigate presence of possible outliers, we will create box plots for the same set-up.
```{r}
p1 <- ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + theme(legend.position = c(0.95,0.4))
p5 <- p4 + scale_x_discrete(name = "Time")
p6 <- p5 + scale_fill_grey(start = 0.5, end = 1)
p6
```
\  

It seems that we can spot multiple potential outliers. These most likely reflect the few deviating individuals within groups that were shown in the earlier plot where rats were individually plotted across time.
\  

Since we are going to summarize the groups, we want to get rid of any outliers within groups. Next we will filter out the above seen outliers and show box plots of mean summary measures for the experimental groups (first with the outliers, then without the outliers).
```{r}
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()

glimpse(RATSL8S)

p1 <- ggplot(RATSL8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time 1-64")
p5

RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
RATSL8S1 <- RATSL8S %>%
  filter(mean > 240)
RATSL8SG3 <- subset(RATSL8S, Group==3, select=c(Group, ID, mean)) %>%
  filter(mean > 500)

RATSL8SG12 <- subset(RATSL8S1, Group==1:2, select=c(Group, ID, mean))

RATSL8S1 <- rbind(RATSL8SG12, RATSL8SG3)

glimpse(RATSL8S1)

p1 <- ggplot(RATSL8S1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight), Time 1-64")
p5
```

Now that the outliers have been removed we can see that the groups obviously differ when their mean response is stacked through time. However, it is equally obvious that we can't infer whether the difference is due to the experimental condition or (what is highly more likely) due to differences in the baseline.
\  

To be exact, we will also run a formal test (ANOVA) for the difference between the three experimental groups.
```{r}
fit1 <- lm(mean ~ Group, data = RATSL8S1)
summary(fit1)
anova(fit1)
```
The test confirms what we inspected earlier from the plot, the group means significantly differ from each other (at least when contrasted to group 1).
\  

Let's do one more model were we include the baseline as a covariate to the model (to be precise, we will use first measurement point after the experiment has started as our baseline measure).
```{r}
baseline <- RATS$WD1
RATSL8S2 <- RATSL8S %>%
  mutate(baseline)

fit2 <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit2)
anova(fit2)
```
When weight in baseline is kept constant in the model we can see that the group differences are not significant anymore (alpha level >.05). Thus, as we suspected, when the baseline weight is controlled for the experimental condition does not have a significant effect for weight. However, since we stacked the individual responses through time and analysed only stacked group means, we should apply some caution before jumping into conclusions, since we are not taking properly into account the repeated measure nature of the data (also, we only have 9 participants in our current sample).
\  

Next we will "Implement the analyses of Chapter 9 of MABS using the BPRS data."
\  

\  

#### **Comparing long and wide form of BPRS**:
For variable description see [https://github.com/ohwhalme/IODS-project/blob/master/data/meet_and_repeat.R](https://github.com/ohwhalme/IODS-project/blob/master/data/meet_and_repeat.R))
```{r}
glimpse(BPRS)
glimpse(BPRSL)
head(BPRSL) 
tail(BPRSL)
```
As was the case with the RATS data, wide form of BPRS contains participant (40) as single rows and the long form stacks participants on top of each other as many times as there are measurement points in the data set (9 * 40 = 360). Wide form contains separate variables for each measurement point (Week0-8) and long form one variable were week number is coded as factor. Long form also includes separate variable for score on bprs.
\  

Next we will visualize the long form of the bprs data.
```{r}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject))
p2 <- p1 + geom_text(aes(label = treatment))
p3 <- p2 + scale_x_continuous(name = "week", breaks = seq(0:8))
p4 <- p3 + scale_y_continuous(name = "bprs")
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```
\  

Here, we have plotted bprs score on y-axis and week number in x-axis. Treatment group is used as a label for each observation. The labels are overlapping so it's quite difficult to interpret what is happening in the plot, but it seems that group 2 is more represented in the high and low ends of the bprs scale throughout the trial compared to group 1.
\  

Next we will fit a linear regression model to the data. We will do this, even though we know that this model is not appropriate fit to the data (due to the repeated measure nature of the data bprs measures are not independent of each other), to demonstrate how the model selection influences the results you receive.
\  

\  

#### **Fitting linear regression model to BPRS**:
```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```
For now, we can say that when week and treatment are included to the model at the same time, only week have statistically significant effect (p<.05) on bprs, where bprs score decreases 2.2 units when week increases one unit.
\  

We will get back to these results after we have calculated the more appropriate linear mixed effects model for the same data.
\  

\  

#### **Fitting linear mixed effects model to BPRS**:
Before running the LME, we will visualize once more the BPRS data set.
```{r}
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ treatment, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
p6

pairs(BPRS, cex = 0.7)
```
\  

**Did not have time for comments :(**
\  

**Model1**
```{r}
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```
\  

**Model2**
```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```
\  

**Model3**
```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref1, BPRS_ref2)
```
**Quick interpretation**: model 1 is LME with random intercept only, 2 with random slope and intercept, 3 with random slope and intercept + interaction for week*treatment. AIC is best for model 3 but ANOVA did not indicate significant difference in fit with Model2 & 3. CIs or p-values for estimates were not provided, but the size of the t-value indicates that the treatment groups did not differ from each other in regard to bprs in the model 2.

